Natural selection in contemporary humans is driven by correlations with human capital

### 2021-04-13 why-natural-selection.Rmd
responses to PNAS review

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
bias could also increase with age.

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization] To learn more about the underlying
mechanisms, we split the sample, starting with basic demographic variables
including education, income and sex. These are all potential sources of
ascertainment bias: as well as the ascertainment for income and education,
mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.

@@

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income. Each point represents a single bivariate regression of number of children on a polygenic score.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

\FloatBarrier
## Correcting for sampling bias

Since UK Biobank participants are not representative of the wider population,
our results suggest that naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables. Since the UK Biobank
sample seems to be ascertained in ways that shrink estimates of selection
effects, we suspect that our weighted estimates are still conservative.[^purifying]

```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions. Unweighted estimates are shown for comparison."}

@@

An alternative theory is that polygenic scores correlate with the
motivation to have children (parameter $a$ in the model). However, this theory
would not explain why our results are weaker at higher incomes and education levels.
Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$, this
relationship actually becomes stronger at higher wages.

The above does not provide a complete theory of fertility. Rather,
it shows that a relatively simple economic model can explain many
of our results. In the appendix, we discuss other models of fertility
from the economic literature, as well as causal evidence for the relationship
between human capital, income and fertility.

This theory can explain why selection effects have opposite signs when AFLB is
controlled for. Suppose that people are aware of their endowments of human
capital, i.e. skills and characteristics that are valuable in the labour market.
Suppose also that education and innate human capital are complements.
Individuals choose how long to stay in education; after leaving education they
enter the labour market and form families. If so, then people with less human
capital will leave education earlier and have their first child earlier [cf.
@caucutt2002women; @monstad2008education]. This is a pure substitution effect:
the income effect plays no role in the decision to leave education, since people
are not yet earning wages.[^no-borrow] Whilst in the labour market, higher
earners will have more children (i.e. income effects dominate). These
assumptions will lead to the pattern we observe: earnings-increasing variants
will predict later first births (and fewer children overall), but they will also
predict more children within any given age at first live birth.

[^no-borrow]: This argument assumes that capital market imperfections hinder
people from borrowing against future income.

The theory can also explain why the natural selection pressure is weaker among
higher-income parents. @becker1976child show that if parents care about child
quality as well as quantity, under certain assumptions the income effect will be
stronger at higher income levels; this can lead to a U-shaped relationship
between income and fertility. Empirically, @cohen2013financial find that income
decreases fertility at low income levels but increases it at higher income
levels. If, in our data, at high income levels income and substitution effects
are roughly balanced , then income-linked polygenic scores will be neither
selected for nor against.

Lastly, it is often assumed that the substitution effect will dominate for lone
parents, or those in unstable relationships, since they have less opportunity to
share childcare responsibilities, while the income effect will dominate for
couples who are able to reap the gains from specialization [@becker1981treatise]. Indeed, US fertility
decreases faster with education among single mothers than married mothers
[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
decrease fertility more among single parents than among couples. The same logic 
could explain our results for lifetime number of sexual partners, if this is 
associated with relationship stability.

@@

n_risk_sig <- res_risk_control %>% 
            filter(term == 'f.2040.0.0') %>% 
            {sum(.$p.value <= 0.05/33)}

@@

An alternative theory is that traits selected for are linked to externalizing
behaviour, risk-seeking and low time discount rates, via the channel of early
sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
for this: scores which might plausibly be linked to externalizing behaviour,
like ADHD and (younger) age at smoking initiation, are selected for. However,
this theory is less good at explaining variation in selection across the full
range of scores, including physical measures, e.g. waist-hip ratio and BMI.
Externalizing behaviour also does not explain why selection effects should work
in the opposite direction among older mothers, whereas the economic theory can
explain this via income effects, as described above. We test the alternative
theory directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040). While risk attitude is always
a highly significant predictor of number of children, it has little impact on
the effect size or significance of the polygenic scores. The median ratio of
effect sizes between regressions with and without controls is `r median(ratio)`;
all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. Thus, although risk attitude does
predict number of children, it appears not to mediate selection effects.
Overall, we believe that the economic theory is the most likely explanation.

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk, or are what many
people would see as undesirable to have. For example, most people would probably
prefer to have high educational attainment, a low risk of ADHD and major
depressive disorder, and a low risk of coronary artery disease, but natural
selection is pushing against genes associated with these traits. Potentially,
this could increase the health burden on modern populations, but that depends on
effect sizes. Our results suggest that naïve estimates can be affected by sample
ascertainment bias. This problem may be less serious in surveys which aim to be
representative (as the UK Biobank does not). However, there is still scope for
bias, since not all respondents consent to the collection of genetic data. For
instance, completion rates for genotype data in the US Health and Retirement
Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks of
ascertainment when studying modern natural selection.

@@

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups.
Overall, natural selection increases inequality for 29 out of 33 polygenic
scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
scores are predictive of disease risk, this increase in inequality could
potentially increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism", including speculation that the Fermi paradox may be resolved by
dysgenic selection among alien species [@sarraf2019modernity]. Others are more
sanguine, and argue that natural selection's effects are outweighed by
environmental improvements, such as those underlying (e.g.) the Flynn effect
[@flynn1987massive]. The evidence here may add some nuance to this debate.
Patterns of natural selection have been relatively consistent across the past two
generations, but they are not the outcome of a universal, society-wide phenomenon.
Instead they result from a sum of opposing forces, operating in different
parts of society and pulling in different directions.

@@

drake::loadd(res_ineq)

>>>>

We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
individuals, corrected for ancestry using 100 genetic principal components (see
Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
polygenic scores in the sample by 5-year birth intervals. Several scores show
consistent increases or declines over this 30-year period, of the order of 5% of
a standard deviation. These changes could reflect natural selection within the
UK population, but also emigration, or ascertainment bias within the sample.
Respondents have higher income and are better educated than the UK population,
and they may also differ on other unobserved characteristics
[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
bias could also increase with age. Lastly, the sample sex ratio skews 
`r female_ratio * 100`% female.

@@

The "selection effect", $\beta$, reflects the strength of natural selection
within the sample. In fact, since polygenic scores are normalized, $\beta$ is
the expected polygenic score among children of the sample
[@beauchamp2016genetic].[^normalization]

@@

Since UK Biobank participants are not representative of the wider population,
naïve estimates of natural selection are at risk of
ascertainment bias. To correct for this, we weight participants using population
data. We try three alternative weighting schemes: weighting by geography, age
and presence/absence of a partner; weighting by age and highest educational
qualification; and age, highest qualification, and age at first live birth, for
women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
entire sample, estimated with the three weighting schemes.
Mean effect sizes across all polygenic scores are
increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
further affected by weighting on other demographic variables.[^purifying]

```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}

To understand the reasons for this ascertainment bias, and to learn about the
mechanisms underlying natural selection, we split the sample, starting with
basic demographic variables including education, income and sex. These are all
potential sources of ascertainment bias, as mentioned above.

```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}

@@

\FloatBarrier

@@

```{r, eval = FALSE}
"Other models of the literature

The above does not provide a complete theory of fertility. Rather, it shows that
a relatively simple economic model can explain many of our results. Other
empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
@becker1981treatise.

@@

n_risk_sig <- res_risk_control %>% 
            filter(term == 'f.2040.0.0') %>% 
            {sum(.$p.value <= .Machine$double.eps)}

@@

An alternative theory is that polygenic scores correlate with the motivation to
have children (parameter $a$ in the model; @jones2008fertility). However, this
theory would not explain why our results are weaker at higher incomes and
education levels. Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$,
this relationship actually becomes stronger at higher wages.

Another alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be captured by our risk-aversion 
parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
[@Mills2020.05.06.081273]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
(younger) age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
remain so when controlling for risk attitude. This non-result could simply
reflect the imprecision of the risk aversion measure, which is a single yes/no
question. However, the measure *does* predict the overall number of children,
highly significantly ($p \le 2^{-16}$, `r n_risk_sig`/33 regressions). Given that,
and our large N, we believe that the non-result is real: while risk aversion does
predict fertility in the sample, it is not an important channel for natural 
selection.

@@

Polygenic scores which correlate with high (low) earnings and more (less)
education are being selected against (selected for). In addition, many of the
phenotypes under positive selection are linked to disease risk. Most people
would probably prefer to have high educational attainment, a low risk of ADHD
and major depressive disorder, and a low risk of coronary artery disease, but
natural selection is pushing against genes associated with these traits.
Potentially, this could increase the health burden on modern populations, but
that depends on effect sizes. Our results suggest that naïve estimates can be
affected by sample ascertainment bias. This problem may be less serious in
surveys which aim to be representative (as the UK Biobank does not). However,
there is still scope for bias, since not all respondents consent to the
collection of genetic data. For instance, completion rates for genotype data in
the US Health and Retirement Study were around 80-85% [@hrs2020]. Researchers
should be aware of the risks of ascertainment when studying modern natural
selection.

@@

```{r calc-inequality}
drake::loadd(res_ineq)

median_pct_change <- median(ineq_comp$pct_change)
n_increases <- sum(ineq_comp$pct_change > 0)

Because selection effects are concentrated in lower-income groups, they may also
increase inequality with respect to polygenic scores. For example, Figure
\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
attainment (EA3) among children from households of different income groups. The
blue bars show the actual means, i.e. parents' mean polygenic score weighted by
number of children. The grey bars show the hypothetical means if all households
had equal numbers of children. Natural selection against genes associated with
educational attainment, which lowers the mean, is stronger at the bottom of the
income distribution, and this increases the differences between groups. Overall,
natural selection increases inequality for `r n_increases` out of 33 polygenic
scores, with a median increase of `r median_pct_change`% in the difference
between highest and lowest income groups (Appendix Table
\@ref(tab:tbl-inequality-effects)). These are substantial effects, and since
many polygenic scores are predictive of disease risk, they could potentially
increase health inequalities. In general, the evolutionary history of
anatomically modern humans is related to disease risk [@benton2021influence];
understanding the role of contemporary natural selection may aid in mapping the
genetic architecture of current health disparities.

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
natural selection's effects are outweighed by environmental improvements, such
as those underlying the Flynn effect [@flynn1987massive]. The evidence
here may add some nuance to this debate. Patterns of natural selection have been
relatively consistent across the past two generations, but they are not the
outcome of a universal, society-wide phenomenon. Instead they result from a sum
of opposing forces, operating in different parts of society and pulling in
different directions.

@@
